A tutorial of Bayesian beta regressions with brms in R

### 2025-02-07 manuscript/manuscript.qmd
edit manuscript

Phonetic research often involves numeric continuous outcome variables, like durations, frequencies, loudness and ratios [@kluender1988; @johnson2003; @gick2013; @ladefoged2014; @abramson2017; @roettger2018; @coretta2023]. Another commonly employed type of outcome variable are proportions: for example, proportion of voicing during closure [@davidson2016], vocal folds contact quotient [@herbst2017], gesture amplitude [@carignan2021], nasalance [@carignan2021]. Moreover, virtually any measure can be MIN-MAX normalised, a procedure which transforms values so that they are in the range 0--1.

@@

We will use brms to fit a Bayesian beta regression [@burkner2017]. The model has voicing proportion as the outcome variable and the following terms: an interaction between vowel (/i, e, a, o, u/) and second consonant C2 (/k, t, p/), centred speech rate (number of syllables per second); as varying (aka random) terms, we included by-speaker varying coefficients for the vowel/consonant interaction and for centred speech rate.[^1] The categorical predictors vowel and C2 are coded using indexing rather than the traditional R contrasts: in R, this corresponds to suppressing the model's intercept with the `0 +` syntax; using indexing instead of contrasts makes it easier to specify priors and uncertainty is distributed equally across all levels of categorical predictors. For pedagogical simplicity, the model will use the default priors as explained in @sec-bayes.

@@

Turning now to the coefficients for vowel and C2, given the indexing approach of coding these variables the model summary and the output of `fixef()` reports the *predictions* in log-odds for each combination of vowel and C2, rather than differences between levels. The CrIs of the vowel/C2 coefficients span all negative log-odds values: these correspond to proportions that are lower than 0.5 (which is 0 in log-odds). This matches the general trends in the raw data, which we plotted in @fig-ita-egg.

Next, we will plot the predicted proportions of each vowel/C2 combination at mean speech rate (i.e. centred speech rate = 0) and then calculate the average pair-wise difference in voicing proportion between /k, t, p/. Finally, we will assess whether there is greater between-vowel variability in /t, p/ relative to /k/.

Before being able to plot the predictions, it's important to get familiar with the so-called posterior draws. \[Bayesian MCMC XXX\]. Posterior draws can be conveniently obtained with the `as_draws_df()`. For the moment, we will extract only the draws of the constant regression coefficients (model variables starting with `b_`). To check which coefficients are available in a model, use `get_variables()` from the tidybayes package. `as_draws_df()` returns a tibble where each column contains the drawn values of a coefficient. @tbl-voi-prop-draws shows the first ten rows and first five columns of the output of `as_draws_df()`. The probability distribution of the drawn values of each coefficient is the posterior probability distribution of that coefficient. Note that, due to the indexing coding of vowel and C2, all coefficient except `b_speech_rate_c` are *predicted log-odds* for each vowel/C2 combination (the drawn values for `b_speech_rate_c` are drawn *differences* in log-odds for each unit increase of speech rate). The drawn values are in log-odds, but we can convert them to proportions with `plogis()` (we will do this when plotting below).

@@

For plotting, we can use ggplot2 statistics layers from the ggdist package. `stat_halfeye()` plots the density of the posterior probability (in grey), its median (point) and CrIs (lines). Let's use 60 and 80% CrIs and transform the log-odds values to proportions with `plogis()`. See @fig-voi-prop-bm-1.

@@

voi_prop_bm_draws_long |> 
  ggplot(aes(plogis(value), vowel)) +
  stat_halfeye(.width = c(0.6, 0.8)) +
  facet_grid(rows = vars(c2))
```

@@

```{r}
#| label: fig-voi-pbdl-c2
#| fig-cap: "Expected predictions of voicing proportion by consonant, averaged across vowels."
voi_prop_bm_draws_long_c2 |> 
  ggplot(aes(plogis(value_mean), c2)) +
  stat_halfeye(.width = c(0.6, 0.8))
```

@@

Once we have the posterior difference, we can obtain CrIs of the difference using `quantile2()` from the posterior package. Beware that the values of the difference are in log-odds! We can convert these into odd-ratios with `exp()`. The output of the code below is shown in @tbl-voi-bm-quant. Odd-ratios indicate the ratio of the difference between A and B, so that 1 means no difference, values greater than 1 indicate an increase in A relative to B and values lower than 1 indicate a decrease in A relative to B. odd-ratios are useful when looking at differences that are in log-odds because while the relative magnitude of the difference in proportion between two groups is the same independent of the baseline proportion, the *absolute* magnitude of the difference depends on the baseline value. For example, an odd-ratio difference of 1.25 would correspond to a proportion increase of 13 percentage points if the baseline proportion is 0.62 but it would correspond to a proportion increase of 17 percentage points if the baseline proportion is 0.73. Of course, in real research contexts it is still usefull to think about absolute magnitudes and their relevance from a conceptual and methodological perspective. In this tutorial we just focus on odd-ratios for simplicity.

@@

Finally, the package marginaleffects \[XXX\] has two other convenience functions that return CrIs of comparisons across predictor levels (`avg_comparisons()`) and CrIs of posterior predictions across predictor levels (`avg_predictions`). \[XXX\]

@@

For the second case study we will use data from @carignan2021. The study looked at properties of nasality in German VNC sequences. Here, we will focus on the effect of C voicing (voiceless /t/ vs voiced /d/) on the proportion of nasalisation within the vowel in the VNC sequence. Previous work on coarticulatory nasalisation in English has suggested that vowels followed by an NC sequence where C is voiceless (NT) should show earlier coarticulatory nasalisation than vowels followed by an NC sequence where C is voiced [ND, see review in @carignan2021]. This pattern has been suggested to be driven by the articulatory and acoustic incompatibility of voicelessness and nasalisation, by which the velum opening gesture of the nasal consonant is pushed away (i.e. earlier) when the consonant following the nasal is voiceless. Everything else being equal, a greater proportion of vowel nasalisation (from the perspective of time) should be found in vowels followed by NT than in vowels followed by ND.

@@

-   `NC` is the NC sequence.

@@

Negative log-odds indicate a proportion that is smaller than 50%, while positive log-odds a proportion that is greater than 50%. Generally, the expected log-odd predictions are negative, indicating an overall tendency for the nasalisation to take less than 50% of the duration of the vowel. Moreover, the predictions are higher for voiceless NC sequences than for voiced NC sequences, indicating a greater proportion of nasalisation in the former. However there is vowel-specific variation, and there doesn't seem to be much of a difference in nasalisation proportion in /e/ and /u/. @fig-nas-prop shows the expected predictions with `conditional_effects()`, which should make the interpretation clearer.

@@

Now that we fitted the model we can use the draws to answer the two research questions (repeated from above):

@@

To answer question 1, we can calculate the average difference in nasalisation proportion by first calculating the average nasalisation across all vowels for voiced and voiceless sequences (see @tbl-nas-draws for the output of this step) and then take the difference of those, similarly to what we have done in the Case Study 1 above.

@@

Now let's calculate the mean nasalisation proportion within each draw by voicing, and plot the resulting posterior distributions. Note that, as discussed for Case study 1, when working wit log-odds it is important to first do all necessary calculations in log-odds, here calculate the mean log-odds across vowels, and *then* transform the calculated estimands to proportions/probabilities. The output of the following code is show in @tbl-nas-long-voi and the density plot of the calculated draws is in @fig-nas-prop-bm-draws-long-voicing-plot.

@@

The plot suggests an overall greater nasalisation proportion in voiceless NC sequences. Let's quantify how greater as we did in Case Study 1. We will use odd-ratios in this context as well, i.e. we will convert log-odds to odd-ratios using the `exp()` (exponential) function (and as before we first calculate the difference and then exponentiate the resulting values, after which we can take summary measures, like means and quantile-based measures such as CrIs). tbl-nas-prop-bm-diff-quant show the 90, 80 and 60% CrIs of the difference ratio of nasalisation proportion in voiceless vs voiced NC sequences.

@@

Moving onto question 2: is there individual speaker variation? \[TAMMINGA XXX\] For this, we will use the `spread_draws()` function from tidybayes \[XXX\] to extract the draws of the varying terms (in brms these are the coefficients that start with `r_`). There is quite a few steps of processing to get from the raw draws to the estimand we need: while we have commented the following code, we encourage readers to test each line sequentialy and inspect the intermediate output to fully understand the process. We assume that readers are familiar enough with models with varying terms (aka random effects, mixed-effects models). What readers should note is that to obtain the expected predictions of nasalisation proportion for each speaker, the constant terms and the varying terms should be added (since the varying terms indicate the deviation of each speaker from the overall estimate).

@@

>>>>

Phonetic research often involves numeric continuous outcome variables, like durations, frequencies, loudness and ratios [@kluender1988; @johnson2003; @gick2013; @ladefoged2014; @abramson2017; @roettger2018; @coretta2023; @rosen2005]. Another commonly employed type of outcome variable are proportions: for example, proportion of voicing during closure [@davidson2016], vocal folds contact quotient [@herbst2017], gesture amplitude [@carignan2021], nasalance [@carignan2021]. Moreover, virtually any measure can be MIN-MAX normalised, a procedure which transforms values so that they are in the range 0--1.

@@

We will use brms to fit a Bayesian beta regression [@burkner2017]. The model has voicing proportion as the outcome variable and the following terms: an interaction between vowel (/i, e, a, o, u/) and second consonant C2 (/k, t, p/), centred speech rate (number of syllables per second); as varying (aka random) terms, we included by-speaker varying coefficients for the vowel/consonant interaction and for centred speech rate.[^1] The categorical predictors vowel and C2 are coded using indexing rather than the traditional R contrasts, as advocated in @mcelreath2019: in R, this corresponds to suppressing the model's intercept with the `0 +` syntax; using indexing instead of contrasts makes it easier to specify priors and uncertainty is distributed equally across all levels of categorical predictors. For pedagogical simplicity, the model will use the default priors as explained in @sec-bayes.

@@

Turning now to the coefficients for vowel and C2, given the indexing approach of coding these variables the model summary and the output of `fixef()` reports the *predictions* in log-odds for each combination of vowel and C2, rather than differences between levels. The CrIs of the vowel/C2 coefficients span all negative log-odds values: these correspond to proportions that are lower than 0.5 (which is 0 in log-odds). This matches the general trends in the raw data, which we plotted in @fig-ita-egg. Next, we will plot the predicted proportions of each vowel/C2 combination at mean speech rate (i.e. centred speech rate = 0) and then calculate the average pair-wise difference in voicing proportion between /k, t, p/.

Before being able to plot the predictions, it's important to get familiar with the so-called posterior draws. Bayesian regression models are fitted using Markov Chain Monte Carlo (MCMC) methods, as briefly explained in @sec-bayes. The `brm()` function runs three MCMC chains with 2000 iterations each by default. These are sufficient in most cases. The first 1000 iterations of each chain are used for "warmup", where the MCMC algorithm is optimised to find the posterior probability. The second half of the iterations, 1000 per chain, are used for sampling from the posterior probability. Each iteration returns a drawn value for each parameter in the model. The list of values drawn during the Monte Carlo sampling are called the "posterior draws". The posterior draws are used for inference: we can plot them, summarise them, transform them and combine them in any way to answer specific questions.

Posterior draws can be conveniently obtained with the `as_draws_df()` from brms. For the moment, we will extract only the draws of the constant regression coefficients (model variables starting with `b_`). To check which coefficients are available in a model, use `get_variables()` from the tidybayes package [@kay2019]. `as_draws_df()` returns a tibble where each column contains the drawn values of a coefficient. @tbl-voi-prop-draws shows the first ten rows and first five columns of the output of `as_draws_df()`. The probability distribution of the drawn values of each coefficient is the posterior probability distribution of that coefficient. Note that, due to the indexing coding of vowel and C2, all coefficient except `b_speech_rate_c` are *predicted log-odds* for each vowel/C2 combination (the drawn values for `b_speech_rate_c` are drawn *differences* in log-odds for each unit increase of speech rate). The drawn values are in log-odds, but we can convert them to proportions with `plogis()` (we will do this when plotting below).

@@

For plotting, we can use ggplot2 statistics layers from the ggdist package [@kay2023; kay2024]. `stat_halfeye()` plots the density of the posterior probability (in grey), its median (point) and CrIs (lines). Let's use 60 and 80% CrIs and transform the log-odds values to proportions with `plogis()`. See @fig-voi-prop-bm-1.

@@

voi_prop_bm_draws_long |> 
  ggplot(aes(plogis(value), vowel)) +
  stat_halfeye(.width = c(0.6, 0.8)) +
  facet_grid(rows = vars(c2)) +
  labs(
    caption = "The points are the medians and the two lines are 60 and 80% CrIs."
  )
```

@@

```{r}
#| label: fig-voi-pbdl-c2
#| fig-cap: "Expected predictions of voicing proportion by consonant, averaged across vowels."
voi_prop_bm_draws_long_c2 |> 
  ggplot(aes(plogis(value_mean), c2)) +
  stat_halfeye(.width = c(0.6, 0.8)) +
  labs(
    caption = "The points are the medians and the two lines are 60 and 80% CrIs."
  )
```

@@

Once we have the posterior difference, we can obtain CrIs of the difference using `quantile2()` from the posterior package [@burkner2024]. Beware that the values of the difference are in log-odds! We can convert these into odd-ratios with `exp()`. The output of the code below is shown in @tbl-voi-bm-quant. Odd-ratios indicate the ratio of the difference between A and B, so that 1 means no difference, values greater than 1 indicate an increase in A relative to B and values lower than 1 indicate a decrease in A relative to B. Odd-ratios are useful when looking at differences that are in log-odds because while the relative magnitude of the difference in proportion between two groups is the same independent of the baseline proportion, the *absolute* magnitude of the difference depends on the baseline value. For example, an odd-ratio difference of 1.25 would correspond to a proportion increase of about 3 percentage points if the baseline proportion is 0.62 but it would correspond to a proportion increase of 5 percentage points if the baseline proportion is 0.73 and 1 percentage point if the baseline is 0.99. Of course, in real research contexts it is still useful to think about absolute magnitudes and their relevance from a conceptual and methodological perspective. In this tutorial we just focus on odd-ratios for simplicity.

```{r}
#| label: odds
#| include: false
#| eval: false

logs <- qlogis(0.62)
p <- plogis(logs * 1.25)
round(p - 0.62, 2)

logs <- qlogis(0.73)
p <- plogis(logs * 1.25)
round(p - 0.73, 2)

logs <- qlogis(0.99)
p <- plogis(logs * 1.25)
round(p - 0.99, 2)
```

@@

Finally, the package marginaleffects [@arelbundock2025] has two other convenience functions that return CrIs of comparisons across predictor levels (`avg_comparisons()`) and CrIs of posterior predictions across predictor levels (`avg_predictions`). Both functions take the model and a list of variables as arguments. Optionally, you can set the confidence level with the `conf_level` argument (here, we set it to 0.8, or 80%). `avg_comparisons()` allows the user to specify the type of comparison for each variable: here we set `"pairwise"` to calculate pairwise comparisons of each level of `c2`. The output of the two functions is shows in @tbl-marginal-1 and @tbl-marginal-2.

@@

For the second case study we will use data from @carignan2021. The study looked at properties of nasality in German VNC sequences. Here, we will focus on the effect of C voicing (voiceless /t/ vs voiced /d/) on the proportion of nasalisation within the vowel in the VNC sequence. Previous work on coarticulatory nasalisation in English has suggested that vowels followed by an NC sequence where C is voiceless (NT) should show earlier coarticulatory nasalisation than vowels followed by an NC sequence where C is voiced (ND). This pattern has been suggested to be driven by the articulatory and acoustic incompatibility of voicelessness and nasalisation, by which the velum opening gesture of the nasal consonant is pushed away (i.e. earlier) when the consonant following the nasal is voiceless [@fowler2005; @beddor2009; @cho2017; @carignan2015]. Everything else being equal, a greater proportion of vowel nasalisation (from the perspective of time) should be found in vowels followed by NT than in vowels followed by ND.

@@

-   `NC` is the NC sequence (voiceless /nt/ or voiced /nd/).

@@

Negative log-odds indicate a proportion that is smaller than 50%, while positive log-odds a proportion that is greater than 50%. Generally, the expected log-odd predictions in @tbl-nas-prop-fixef are negative, indicating an overall tendency for the nasalisation to take less than 50% of the duration of the vowel. Moreover, the predictions are higher for voiceless NC sequences than for voiced NC sequences, indicating a greater proportion of nasalisation in the former. However there is vowel-specific variation, and there doesn't seem to be much of a difference in nasalisation proportion in /e/ and /u/. @fig-nas-prop shows the expected predictions with `conditional_effects()`, which should make the interpretation of the predictions from @tbl-nas-prop-fixef clearer.

@@

Now that we fitted the model we can use the posterior draws to answer the two research questions (repeated from above):

@@

To answer question 1, we can calculate the average difference in nasalisation proportion by first calculating the average nasalisation across all vowels for voiced and voiceless sequences (see @tbl-nas-draws for the output of this step) and then take the difference of those, similarly to what we have done in the case study in @sec-case-1.

@@

Now let's calculate the mean nasalisation proportion within each draw by voicing, and plot the resulting posterior distributions. Note that, as discussed in @sec-case-1, when working with log-odds it is important to first do all necessary calculations in log-odds (here calculate the mean log-odds across vowels) and *then* transform the calculated estimands to proportions/probabilities. The output of the following code is show in @tbl-nas-long-voi and the density plot of the calculated draws is in @fig-nas-prop-bm-draws-long-voicing-plot.

@@

The plot suggests an overall greater nasalisation proportion in voiceless NC sequences. Let's quantify how greater as we did in @sec-case-1. We will use odd-ratios in this context as well, i.e. we will convert log-odds to odd-ratios using the `exp()` (exponential) function (and as before we first calculate the difference and then exponentiate the resulting values, after which we can take summary measures, like means and quantile-based measures such as CrIs). @tbl-nas-prop-bm-diff-quant show the 90, 80 and 60% CrIs of the difference ratio of nasalisation proportion in voiceless vs voiced NC sequences.

@@

Moving onto question 2: is there individual speaker variation? While in the recent tradition of linguistic research to use varying (aka random) effects to *control* for differences across participants, it has been proposed to use the estimates of the varying coefficients to investigate individual differences [@tamminga2016]. We will use the `spread_draws()` function from tidybayes [@kay2019] to extract the draws of the varying terms (in brms these are the coefficients that start with `r_`). There is quite a few steps of processing to get from the raw draws to the estimand we need: while we have commented the following code, we encourage readers to test each line sequentially and inspect the intermediate output to fully understand the process. We assume that readers are familiar enough with models with varying terms (aka random effects, mixed-effects models). What readers should note is that to obtain the expected predictions of nasalisation proportion for each speaker, the constant terms and the varying terms should be added (since the varying terms indicate the deviation of each speaker from the overall estimate).

@@

# Conclusion

In this tutorial we have presented two case studies where the outcome variable is a beta variable, i.e. a numeric continuous variable bounded between 0 and 1. We have shown how to fit Bayesian beta regressions to coarticulatory progressive voicing in voiceless stops in Italian in @sec-case-1 and coarticulatory regressive nasalisation in vowels in German in @sec-case-2. While this tutorial alone will not be sufficient to be able to independently run full Bayesian analyses, it will serve as a first introduction for readers to jump start their learning.
